---
title: "Simulation Study of Regression Variable Selection"
author: "Mojde Mir"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```




# Packages and folders

```{r packages-folders}
rm(list = ls())

library(hdrm)
library(glmnet)
library(olsrr)
library(dplyr)
library(tidyr)
library(future)
library(future.apply)
library(progressr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(parallel)


Sys.setenv(OMP_NUM_THREADS = 1)
Sys.setenv(MKL_NUM_THREADS = 1)

# Increase allowed size of objects exported to parallel workers
options(future.globals.maxSize = 8 * 1024^3)

project_folder <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project4"

table_folder <- file.path(project_folder, "Results", "Tables")
figure_folder <- file.path(project_folder, "Results", "Figures")
```

# Design choices

```{r design-settings}
set.seed(6624)

B <- 10000

num_predictors <- 20
num_signal <- 5
predictor_names <- paste0("X", seq_len(num_predictors))

beta_values <- c(0.5/3, 1/3, 1.5/3, 2/3, 2.5/3, rep(0, 15))
names(beta_values) <- predictor_names

signal_variables <- predictor_names[1:num_signal]
noise_variables <- predictor_names[(num_signal + 1):num_predictors]

sig_level <- 0.05
enet_mixing <- 0.5
folds_for_cv <- 10

design_grid <- expand.grid(
  sample_size = c(250, 500),
  rho = c(0, 0.35, 0.70)
)

design_grid
```

# Helper functions

## Monte Carlo SEs

The Monte Carlo standard error is reported for simulation-based averages and proportions. This is helpful because small differences between methods may be due to simulation noise.

```{r simulation-error-functions}
se_for_average <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) < 2) return(NA_real_)
  sd(x) / sqrt(length(x))
}

se_for_proportion <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  pbar <- mean(x)
  sqrt(pbar * (1 - pbar) / length(x))
}
```

## Generate data

```{r data-generator}
make_regression_sample <- function(sample_size, rho_value, beta = beta_values) {

  generated <- hdrm::gen_data(
    n = sample_size,
    p = num_predictors,
    p1 = num_signal,
    beta = beta,
    family = "gaussian",
    signal = "heterogeneous",
    corr = "exchangeable",
    rho = rho_value
  )

  predictors <- as.data.frame(generated$X)
  colnames(predictors) <- predictor_names

  list(
    response = as.numeric(generated$y),
    predictors = predictors
  )
}
```

## Quick data check

```{r data-check}
example_data <- make_regression_sample(sample_size = 250, rho_value = 0.35)

length(example_data$response)
dim(example_data$predictors)
round(colMeans(example_data$predictors), 3)
round(apply(example_data$predictors, 2, sd), 3)
round(cor(example_data$predictors), 2)
```

## Pull variable names from fitted models

```{r selected-variable-extraction}
terms_from_lm <- function(model_object) {
  setdiff(names(coef(model_object)), "(Intercept)")
}

terms_from_cv_glmnet <- function(cv_object, lambda_label) {
  coefficient_vector <- coef(cv_object, s = lambda_label)
  kept_terms <- rownames(coefficient_vector)[as.numeric(coefficient_vector) != 0]
  setdiff(kept_terms, "(Intercept)")
}
```

# Variable-selection procedures

This function applies each selection rule to one simulated data set. It only returns the selected variable names; estimation and inference are handled afterward in a separate step.

```{r run-selection-methods}
apply_selection_rules <- function(response, predictors, sample_size) {

  modeling_data <- data.frame(Y = response, predictors)
  starting_model <- lm(Y ~ ., data = modeling_data)

  # Backward removal by p-value.
  pvalue_step <- ols_step_backward_p(
    starting_model,
    prem = sig_level,
    details = FALSE
  )

  pvalue_vars <- terms_from_lm(pvalue_step$model)

  # Backward AIC. In step(), the default k = 2 gives AIC.
  aic_model <- stats::step(
    starting_model,
    direction = "backward",
    trace = 0
  )

  aic_vars <- terms_from_lm(aic_model)

  # Backward BIC. Setting k = log(n) changes the penalty from AIC to BIC.
  bic_model <- stats::step(
    starting_model,
    direction = "backward",
    k = log(sample_size),
    trace = 0
  )

  bic_vars <- terms_from_lm(bic_model)

  x_matrix <- as.matrix(predictors)

  lasso_fit <- cv.glmnet(
    x = x_matrix,
    y = response,
    alpha = 1,
    nfolds = folds_for_cv,
    standardize = TRUE
  )

  elastic_fit <- cv.glmnet(
    x = x_matrix,
    y = response,
    alpha = enet_mixing,
    nfolds = folds_for_cv,
    standardize = TRUE
  )

  list(
    "Backward p-value" = pvalue_vars,
    "Backward AIC" = aic_vars,
    "Backward BIC" = bic_vars,
    "LASSO: lambda.min" = terms_from_cv_glmnet(lasso_fit, "lambda.min"),
    "LASSO: lambda.1se" = terms_from_cv_glmnet(lasso_fit, "lambda.1se"),
    "Elastic net: lambda.min" = terms_from_cv_glmnet(elastic_fit, "lambda.min"),
    "Elastic net: lambda.1se" = terms_from_cv_glmnet(elastic_fit, "lambda.1se")
  )
}
```

# Post-selection OLS step

For each selected model, OLS is refit using only the selected predictors. If no predictors are selected, the function returns an empty data frame and the later evaluation step treats all variables as unselected.

```{r selected-model-refit}
fit_ols_after_selection <- function(response, predictors, kept_variables) {

  kept_variables <- intersect(kept_variables, colnames(predictors))

  if (length(kept_variables) == 0) {
    return(data.frame(
      variable = character(),
      beta_hat = numeric(),
      p_value = numeric(),
      lower_ci = numeric(),
      upper_ci = numeric()
    ))
  }

  modeling_data <- data.frame(Y = response, predictors)

  final_model <- lm(
    reformulate(kept_variables, response = "Y"),
    data = modeling_data
  )

  coefficient_table <- summary(final_model)$coefficients
  confidence_table <- suppressMessages(confint(final_model))

  data.frame(
    variable = rownames(coefficient_table)[-1],
    beta_hat = coefficient_table[-1, "Estimate"],
    p_value = coefficient_table[-1, "Pr(>|t|)"],
    lower_ci = confidence_table[-1, 1],
    upper_ci = confidence_table[-1, 2],
    row.names = NULL
  )
}
```

# Evaluate one selected model

The evaluation table keeps one row for each possible predictor. This makes it easier to compute true positive rates, false positive rates, bias, and coverage across repetitions.

```{r evaluate-one-method}
score_selected_model <- function(method_name, kept_variables, response, predictors,
                                 sample_size, rho_value, replicate_id) {

  ols_table <- fit_ols_after_selection(
    response = response,
    predictors = predictors,
    kept_variables = kept_variables
  )

  variable_level <- data.frame(
    sample_size = sample_size,
    rho = rho_value,
    replicate = replicate_id,
    method = method_name,
    variable = names(beta_values),
    beta_true = as.numeric(beta_values),
    stringsAsFactors = FALSE
  ) %>%
    mutate(
      chosen = variable %in% kept_variables,
      is_signal = beta_true != 0,
      is_noise = beta_true == 0
    ) %>%
    left_join(ols_table, by = "variable") %>%
    mutate(
      beta_unconditional = ifelse(chosen, beta_hat, 0),
      unconditional_bias = beta_unconditional - beta_true,

      significant_after_refit = ifelse(chosen, p_value < sig_level, FALSE),

      ci_covers_truth = case_when(
        chosen ~ lower_ci <= beta_true & beta_true <= upper_ci,
        !chosen & is_noise ~ TRUE,
        !chosen & is_signal ~ FALSE
      ),

      type_I_error = is_noise & chosen & significant_after_refit,
      type_II_error = is_signal & (!chosen | !significant_after_refit)
    )

  selected_signal <- sum(variable_level$chosen & variable_level$is_signal)
  selected_noise <- sum(variable_level$chosen & variable_level$is_noise)
  missed_signal <- sum(!variable_level$chosen & variable_level$is_signal)
  rejected_noise <- sum(!variable_level$chosen & variable_level$is_noise)

  model_level <- data.frame(
    sample_size = sample_size,
    rho = rho_value,
    replicate = replicate_id,
    method = method_name,

    # Confusion-matrix counts for variable selection.
    # TP: true signal selected
    # FP: noise variable selected
    # FN: true signal missed
    # TN: noise variable correctly excluded
    TP = selected_signal,
    FP = selected_noise,
    FN = missed_signal,
    TN = rejected_noise,

    model_size = selected_signal + selected_noise,
    exact_recovery = selected_signal == num_signal & selected_noise == 0,

    true_positive_rate = selected_signal / num_signal,
    false_positive_rate = selected_noise / length(noise_variables),
    true_negative_rate = rejected_noise / length(noise_variables),
    false_negative_rate = missed_signal / num_signal
  )

  list(
    variable_level = variable_level,
    model_level = model_level
  )
}
```

# Run one simulation replicate

```{r one-replicate-function}
run_one_replicate <- function(sample_size, rho_value, replicate_id) {

  sim_data <- make_regression_sample(
    sample_size = sample_size,
    rho_value = rho_value
  )

  selected_sets <- apply_selection_rules(
    response = sim_data$response,
    predictors = sim_data$predictors,
    sample_size = sample_size
  )

  replicate_results <- lapply(names(selected_sets), function(this_method) {
    score_selected_model(
      method_name = this_method,
      kept_variables = selected_sets[[this_method]],
      response = sim_data$response,
      predictors = sim_data$predictors,
      sample_size = sample_size,
      rho_value = rho_value,
      replicate_id = replicate_id
    )
  })

  list(
    variable_level = bind_rows(lapply(replicate_results, `[[`, "variable_level")),
    model_level = bind_rows(lapply(replicate_results, `[[`, "model_level"))
  )
}
```

# Check the workflow

```{r test-workflow}
trial_run <- run_one_replicate(sample_size = 250, rho_value = 0, replicate_id = 1)

head(trial_run$variable_level)
head(trial_run$model_level)
```

# Full simulation

The original parallel version used `future_lapply()`. On some computers, especially when `cv.glmnet()` is also doing repeated cross-validation, the parallel workers can stop before the simulation begins. This version runs the simulation in parallel chunks using `future_lapply()` and saves checkpoints after each chunk.


```{r full-simulation}
B <- 10000

num_workers <- max(1, floor(parallel::detectCores() * 0.75))

run_plan <- tidyr::expand_grid(
  design_id = seq_len(nrow(design_grid)),
  replicate = seq_len(B)
) %>%
  mutate(
    sample_size = design_grid$sample_size[design_id],
    rho = design_grid$rho[design_id]
  ) %>%
  select(sample_size, rho, replicate)

nrow(run_plan)

checkpoint_folder <- file.path(project_folder, "Results", "Checkpoints")
dir.create(checkpoint_folder, recursive = TRUE, showWarnings = FALSE)

chunk_size <- 500
run_plan$chunk_id <- ceiling(seq_len(nrow(run_plan)) / chunk_size)

plan(multisession, workers = num_workers)

clock_start <- Sys.time()

for (this_chunk in sort(unique(run_plan$chunk_id))) {

  checkpoint_file <- file.path(
    checkpoint_folder,
    paste0("simulation_chunk_", this_chunk, ".rds")
  )

  if (file.exists(checkpoint_file)) {
    message("Skipping completed chunk ", this_chunk)
    next
  }

  chunk_plan <- run_plan %>%
    filter(chunk_id == this_chunk)

  message(
    "Starting chunk ", this_chunk, " of ",
    max(run_plan$chunk_id),
    " with ", nrow(chunk_plan), " jobs"
  )

 chunk_results <- future.apply::future_lapply(
  seq_len(nrow(chunk_plan)),
  future.seed = TRUE,
  future.packages = c(
    "hdrm", "glmnet", "olsrr", "dplyr", "tidyr"
  ),
  function(row_id) {

    tryCatch({

      run_one_replicate(
        sample_size = chunk_plan$sample_size[row_id],
        rho_value = chunk_plan$rho[row_id],
        replicate_id = chunk_plan$replicate[row_id]
      )

    }, error = function(e) {

      message(
        "Error in replicate ",
        chunk_plan$replicate[row_id],
        ": ",
        conditionMessage(e)
      )

      NULL
    })
  }
)
  saveRDS(chunk_results, checkpoint_file)

  message("Saved chunk ", this_chunk)
  gc()
}

clock_end <- Sys.time()

message(
  "Simulation time in minutes: ",
  round(as.numeric(difftime(clock_end, clock_start, units = "mins")), 2)
)
```

# Save combined simulation results

```{r save-combined-results}
checkpoint_files <- list.files(
  checkpoint_folder,
  pattern = "^simulation_chunk_.*\\.rds$",
  full.names = TRUE
)

raw_simulation_results <- unlist(
  lapply(checkpoint_files, readRDS),
  recursive = FALSE
)

raw_simulation_results <- raw_simulation_results[
  !sapply(raw_simulation_results, is.null)
]

by_variable <- bind_rows(
  lapply(raw_simulation_results, function(x) x$variable_level)
)

by_model <- bind_rows(
  lapply(raw_simulation_results, function(x) x$model_level)
)

saveRDS(by_variable, file.path(table_folder, "sim_results_by_variable.rds"))
saveRDS(by_model, file.path(table_folder, "sim_results_by_model.rds"))

write.csv(by_variable, file.path(table_folder, "sim_results_by_variable.csv"), row.names = FALSE)
write.csv(by_model, file.path(table_folder, "sim_results_by_model.csv"), row.names = FALSE)
```

# Summary tables

## Method-level performance

This table gives the average TP, FP, FN, and TN counts for each combination of sample size, correlation, and method. The rates are also included to make the counts easier to compare across methods.

```{r method-performance-summary}
method_performance <- by_model %>%
  group_by(sample_size, rho, method) %>%
  summarize(
    mean_TP = mean(TP),
    mean_FP = mean(FP),
    mean_FN = mean(FN),
    mean_TN = mean(TN),

    mean_model_size = mean(model_size),
    mcse_model_size = se_for_average(model_size),

    mean_true_positive_rate = mean(true_positive_rate),
    mcse_true_positive_rate = se_for_average(true_positive_rate),

    mean_false_positive_rate = mean(false_positive_rate),
    mcse_false_positive_rate = se_for_average(false_positive_rate),

    mean_true_negative_rate = mean(true_negative_rate),
    mcse_true_negative_rate = se_for_average(true_negative_rate),

    mean_false_negative_rate = mean(false_negative_rate),
    mcse_false_negative_rate = se_for_average(false_negative_rate),

    exact_model_recovery = mean(exact_recovery),
    mcse_exact_model_recovery = se_for_proportion(exact_recovery),

    .groups = "drop"
  )

write.csv(method_performance, file.path(table_folder, "method_level_TP_FP_FN_TN.csv"), row.names = FALSE)

kable(method_performance, digits = 3)
```



## True positive and Type II error summary for signal variables


```{r appendix-table-A1}
appendix_A1 <- by_variable %>%
  filter(variable %in% signal_variables) %>%
  group_by(sample_size, rho, method, variable, beta_true) %>%
  summarize(
    selection_probability = mean(chosen),
    mean_bias = mean(unconditional_bias, na.rm = TRUE),
    coverage = mean(ci_covers_truth, na.rm = TRUE),
    type_II_error = mean(type_II_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(sample_size, rho, method, variable)

write.csv(
  appendix_A1,
  file.path(table_folder, "Appendix_Table_A1.csv"),
  row.names = FALSE
)

appendix_A1 %>%
  rename(
    N = sample_size,
    rho = rho,
    Method = method,
    Variable = variable,
    `True Beta` = beta_true,
    `Selection Probability` = selection_probability,
    `Mean Bias` = mean_bias,
    Coverage = coverage,
    `Type II Error` = type_II_error
  ) %>%
  kable(
    digits = 4,
    caption = "Appendix Table A1. Variable-Level Performance for Truly Associated Predictors",
    longtable = TRUE,
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("repeat_header"),
    font_size = 9
  )
```

## False positive and Type I error summary for noise variables


```{r appendix-table-A2}
appendix_A2 <- by_variable %>%
  filter(variable %in% noise_variables) %>%
  group_by(sample_size, rho, method, variable) %>%
  summarize(
    false_positive_rate = mean(chosen),
    type_I_error = mean(type_I_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(sample_size, rho, method, variable)

write.csv(
  appendix_A2,
  file.path(table_folder, "Appendix_Table_A2.csv"),
  row.names = FALSE
)

appendix_A2 %>%
  rename(
    N = sample_size,
    rho = rho,
    Method = method,
    Variable = variable,
    `False Positive Rate` = false_positive_rate,
    `Type I Error` = type_I_error
  ) %>%
  kable(
    digits = 4,
    caption = "Appendix Table A2. Variable-Level False Positive Rates and Type I Error for Null Predictors",
    longtable = TRUE,
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("repeat_header"),
    font_size = 9
  )
```

## Bias and coverage summary for signal variables

```{r bias-coverage-summary}
bias_coverage_summary <- by_variable %>%
  filter(variable %in% signal_variables) %>%
  group_by(sample_size, rho, method, variable, beta_true) %>%
  summarize(
    mean_beta_unconditional = mean(beta_unconditional, na.rm = TRUE),
    mean_unconditional_bias = mean(unconditional_bias, na.rm = TRUE),
    mcse_unconditional_bias = se_for_average(unconditional_bias),
    coverage_probability = mean(ci_covers_truth, na.rm = TRUE),
    mcse_coverage_probability = se_for_proportion(ci_covers_truth),
    .groups = "drop"
  )

write.csv(bias_coverage_summary, file.path(table_folder, "bias_coverage_signal_variables.csv"), row.names = FALSE)

kable(bias_coverage_summary, digits = 3)
```

## Overall inference summary

```{r overall-inference-summary}
overall_inference_summary <- by_variable %>%
  group_by(sample_size, rho, method) %>%
  summarize(
    mean_type_I_error = mean(type_I_error[beta_true == 0], na.rm = TRUE),
    mcse_type_I_error = se_for_proportion(type_I_error[beta_true == 0]),

    mean_type_II_error = mean(type_II_error[beta_true != 0], na.rm = TRUE),
    mcse_type_II_error = se_for_proportion(type_II_error[beta_true != 0]),

    mean_bias_signal_variables = mean(unconditional_bias[beta_true != 0], na.rm = TRUE),
    mean_coverage_signal_variables = mean(ci_covers_truth[beta_true != 0], na.rm = TRUE),
    .groups = "drop"
  )

write.csv(overall_inference_summary, file.path(table_folder, "overall_inference_summary.csv"), row.names = FALSE)

kable(overall_inference_summary, digits = 3)
```
