## Table 1: Power and Type II error for the true predictors

```{r true-variable-table, eval=FALSE}
true_variable_results <- by_variable %>%
  filter(variable %in% signal_variables) %>%
  group_by(sample_size, rho, method, variable, beta_true) %>%
  summarize(
    selection_rate = mean(chosen),
    mcse_selection_rate = se_for_proportion(chosen),
    selection_only_type_II = 1 - selection_rate,
    post_refit_type_II = mean(type_II_error, na.rm = TRUE),
    mcse_post_refit_type_II = se_for_proportion(type_II_error),
    .groups = "drop"
  )

write.csv(true_variable_results, file.path(table_folder, "power_and_type2_true_variables.csv"), row.names = FALSE)

kable(true_variable_results, digits = 3)
```

## Table 2: Noise-variable selection

```{r noise-variable-table, eval=FALSE}
noise_variable_results <- by_variable %>%
  filter(variable %in% noise_variables) %>%
  group_by(sample_size, rho, method, variable) %>%
  summarize(
    false_selection_rate = mean(chosen),
    mcse_false_selection_rate = se_for_proportion(chosen),
    .groups = "drop"
  )

write.csv(noise_variable_results, file.path(table_folder, "false_selection_noise_variables.csv"), row.names = FALSE)

kable(noise_variable_results, digits = 3)
```

## Table 3: Bias and confidence interval coverage for true predictors

Unconditional bias uses an estimate of zero when the variable was not selected. This summarizes the full procedure: selection followed by estimation.

```{r bias-coverage-table, eval=FALSE}
bias_coverage_results <- by_variable %>%
  filter(variable %in% signal_variables) %>%
  group_by(sample_size, rho, method, variable, beta_true) %>%
  summarize(
    avg_unconditional_estimate = mean(beta_unconditional, na.rm = TRUE),
    avg_unconditional_bias = mean(unconditional_bias, na.rm = TRUE),
    mcse_unconditional_bias = se_for_average(unconditional_bias),
    ci_coverage = mean(ci_covers_truth, na.rm = TRUE),
    mcse_ci_coverage = se_for_proportion(ci_covers_truth),
    .groups = "drop"
  )

write.csv(bias_coverage_results, file.path(table_folder, "bias_and_coverage_true_variables.csv"), row.names = FALSE)

kable(bias_coverage_results, digits = 3)
```

## Table 4: Type I error for null predictors

Here, a Type I error means the variable is truly null, was selected, and was significant in the post-selection OLS model.

```{r type1-table, eval=FALSE}
type1_results <- by_variable %>%
  filter(variable %in% noise_variables) %>%
  group_by(sample_size, rho, method, variable) %>%
  summarize(
    type_I_error_rate = mean(type_I_error, na.rm = TRUE),
    mcse_type_I_error_rate = se_for_proportion(type_I_error),
    .groups = "drop"
  )

write.csv(type1_results, file.path(table_folder, "type1_error_noise_variables.csv"), row.names = FALSE)

kable(type1_results, digits = 3)
```

## Appendix table: all variable-level summaries

```{r appendix-variable-summary, eval=FALSE}
variable_summary_all <- by_variable %>%
  group_by(sample_size, rho, method, variable, beta_true) %>%
  summarize(
    selection_rate = mean(chosen),
    mcse_selection_rate = se_for_proportion(chosen),

    avg_unconditional_estimate = mean(beta_unconditional, na.rm = TRUE),
    avg_unconditional_bias = mean(unconditional_bias, na.rm = TRUE),
    mcse_unconditional_bias = se_for_average(unconditional_bias),

    ci_coverage = mean(ci_covers_truth, na.rm = TRUE),
    mcse_ci_coverage = se_for_proportion(ci_covers_truth),

    avg_estimate_among_selected = ifelse(
      any(chosen),
      mean(beta_hat[chosen], na.rm = TRUE),
      NA_real_
    ),

    avg_bias_among_selected = ifelse(
      any(chosen),
      mean((beta_hat - beta_true)[chosen], na.rm = TRUE),
      NA_real_
    ),

    selected_only_coverage = ifelse(
      any(chosen),
      mean((lower_ci <= beta_true & beta_true <= upper_ci)[chosen], na.rm = TRUE),
      NA_real_
    ),

    type_I_error_rate = ifelse(
      first(beta_true) == 0,
      mean(type_I_error, na.rm = TRUE),
      NA_real_
    ),

    type_II_error_rate = ifelse(
      first(beta_true) != 0,
      mean(type_II_error, na.rm = TRUE),
      NA_real_
    ),

    .groups = "drop"
  )

write.csv(variable_summary_all, file.path(table_folder, "appendix_all_variable_summaries.csv"), row.names = FALSE)

kable(variable_summary_all, digits = 3)
```

## Overall inference summary

```{r overall-inference-summary, eval=FALSE}
overall_inference <- variable_summary_all %>%
  group_by(sample_size, rho, method) %>%
  summarize(
    average_type_I_error = mean(type_I_error_rate[beta_true == 0], na.rm = TRUE),
    mcse_average_type_I_error = se_for_average(type_I_error_rate[beta_true == 0]),

    average_type_II_error = mean(type_II_error_rate[beta_true != 0], na.rm = TRUE),
    mcse_average_type_II_error = se_for_average(type_II_error_rate[beta_true != 0]),

    average_bias_for_signal_variables = mean(avg_unconditional_bias[beta_true != 0], na.rm = TRUE),
    average_coverage_for_signal_variables = mean(ci_coverage[beta_true != 0], na.rm = TRUE),

    .groups = "drop"
  )

write.csv(overall_inference, file.path(table_folder, "overall_post_selection_inference.csv"), row.names = FALSE)

kable(overall_inference, digits = 3)
```

# Optional visual summaries

```{r optional-plots, eval=FALSE}
plot_model_size <- method_performance %>%
  ggplot(aes(x = method, y = avg_model_size)) +
  geom_col() +
  facet_grid(sample_size ~ rho, labeller = label_both) +
  coord_flip() +
  labs(
    title = "Average selected model size",
    x = "Selection method",
    y = "Average number of selected predictors"
  )

ggsave(file.path(figure_folder, "average_model_size.png"), plot_model_size, width = 9, height = 6)

plot_exact <- method_performance %>%
  ggplot(aes(x = method, y = exact_model_recovery)) +
  geom_col() +
  facet_grid(sample_size ~ rho, labeller = label_both) +
  coord_flip() +
  labs(
    title = "Exact model recovery rate",
    x = "Selection method",
    y = "Proportion of simulations"
  )

ggsave(file.path(figure_folder, "exact_model_recovery.png"), plot_exact, width = 9, height = 6)

plot_noise <- method_performance %>%
  ggplot(aes(x = method, y = avg_noise_selection_rate)) +
  geom_col() +
  facet_grid(sample_size ~ rho, labeller = label_both) +
  coord_flip() +
  labs(
    title = "Average noise-variable selection rate",
    x = "Selection method",
    y = "Average rate"
  )

ggsave(file.path(figure_folder, "noise_selection_rate.png"), plot_noise, width = 9, height = 6)
```

# Interpretation notes

Several patterns are expected in this simulation. BIC and the `lambda.1se` penalized methods should usually select smaller models, which may reduce false positives but increase missed true predictors. AIC and `lambda.min` may select more variables, which can increase power but also tends to include more noise variables. Larger sample size should improve selection of the true predictors. Stronger predictor correlation can make the selection problem harder because correlated variables may compete with each other.

The exact model recovery rate is included as an additional summary. This is a stricter metric than true positive rate or false positive rate because it requires selecting all five true predictors and none of the fifteen null predictors in the same replicate.

Finally, the confidence interval and p-value results should be interpreted as post-selection summaries. Since the same data are used both to select and refit the model, the usual OLS inference does not fully reflect the uncertainty from the model-selection process.
