

---
title: "Data_Overview"
author: "Moj"
date: "2026-03-16"
output: html_document
---

# ============================================================
# BIOS6624 Project 2 - Script 01: Preliminary Data Overview
#
# Purpose:
#   - Read preliminary dataset
#   - Examine structure and summary statistics
#   - Review plots of variables
#   - Calculate skewness
#   - Estimate correlations among predictors and outcomes
#
# Outputs:
#   Results/Tables/summary_table.csv
#   Results/Tables/skewness_table.csv
#   Results/Tables/correlation_table.csv
#   Results/Figures/*.png
# ============================================================


```{r}
# -----------------------
# Clean environment
# -----------------------
rm(list = ls())

# -----------------------
# Load libraries
# -----------------------
library(tidyverse)
library(readr)
library(e1071)

# -----------------------
# Define directories
# -----------------------
prelim_path <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/DataPrelim/PrelimData.csv"
results_tbl <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/Results/Tables"
results_fig <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/Results/Figures"
output_dir  <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/Output"


# Create folders if they don't exist
dir.create(results_tbl, recursive = TRUE, showWarnings = FALSE)
dir.create(results_fig, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# -----------------------
# Read raw data, blank as NA
# -----------------------
data <- read_csv(prelim_path, na = c("", " ", "NA"))
```




```{r}
# -----------------------
# Inspect structure
# -----------------------
str(data)
summary(data)

# -----------------------
# Create log variables
# -----------------------
data$log_IL6  <- log(data$IL_6 + 0.001)
data$log_MCP1 <- log(data$MCP_1 + 0.001)


# -----------------------
# Summary table
# -----------------------
summary_table <- data.frame(
  variable = c("IL_6", "MCP_1", "CVLT_CNG3", "CORT_CNG3"),
  n = c(sum(!is.na(data$IL_6)),
        sum(!is.na(data$MCP_1)),
        sum(!is.na(data$CVLT_CNG3)),
        sum(!is.na(data$CORT_CNG3))),
  mean = c(mean(data$IL_6, na.rm = TRUE),
           mean(data$MCP_1, na.rm = TRUE),
           mean(data$CVLT_CNG3, na.rm = TRUE),
           mean(data$CORT_CNG3, na.rm = TRUE)),
  sd = c(sd(data$IL_6, na.rm = TRUE),
         sd(data$MCP_1, na.rm = TRUE),
         sd(data$CVLT_CNG3, na.rm = TRUE),
         sd(data$CORT_CNG3, na.rm = TRUE))
)

print(summary_table)
write.csv(summary_table, file.path(results_tbl, "summary_table.csv"), row.names = FALSE)
```


```{r}
# -----------------------
# Skewness table
# -----------------------
skewness_table <- data.frame(
  variable = c("IL_6", "MCP_1", "log_IL6", "log_MCP1", "CVLT_CNG3", "CORT_CNG3"),
  skewness = c(
    skewness(data$IL_6, na.rm = TRUE),
    skewness(data$MCP_1, na.rm = TRUE),
    skewness(data$log_IL6, na.rm = TRUE),
    skewness(data$log_MCP1, na.rm = TRUE),
    skewness(data$CVLT_CNG3, na.rm = TRUE),
    skewness(data$CORT_CNG3, na.rm = TRUE)
  )
)

print(skewness_table)
write.csv(skewness_table, file.path(results_tbl, "skewness_table.csv"), row.names = FALSE)
```


```{r}
# -----------------------
# Combined histograms
# -----------------------

png(file.path(results_fig, "all_histograms.png"), width = 1600, height = 1200, res = 150)

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))

hist(data$IL_6,
     main = "IL-6",
     xlab = "IL-6",
     col = "gray80",
     border = "white")

hist(data$log_IL6,
     main = "log(IL-6)",
     xlab = "log(IL-6)",
     col = "gray80",
     border = "white")

hist(data$MCP_1,
     main = "MCP-1",
     xlab = "MCP-1",
     col = "gray80",
     border = "white")

hist(data$log_MCP1,
     main = "log(MCP-1)",
     xlab = "log(MCP-1)",
     col = "gray80",
     border = "white")

hist(data$CVLT_CNG3,
     main = "CVLT Change",
     xlab = "CVLT_CNG3",
     col = "gray80",
     border = "white")

hist(data$CORT_CNG3,
     main = "Cortical Change",
     xlab = "CORT_CNG3",
     col = "gray80",
     border = "white")

mtext("Histograms of Inflammatory Markers and Outcome Change Scores",
      outer = TRUE, cex = 1.4)

dev.off()


# -----------------------
# Combined scatterplots
# -----------------------

png(file.path(results_fig, "all_scatterplots.png"), width = 2000, height = 1600, res = 150)

par(mfrow = c(2, 4), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))

plot(data$IL_6, data$CVLT_CNG3,
     xlab = "IL-6", ylab = "CVLT Change",
     main = "IL-6 vs CVLT",
     pch = 19, col = "gray40")
abline(lm(CVLT_CNG3 ~ IL_6, data = data), col = "blue", lwd = 2)

plot(data$IL_6, data$CORT_CNG3,
     xlab = "IL-6", ylab = "Cortical Change",
     main = "IL-6 vs Cortical",
     pch = 19, col = "gray40")
abline(lm(CORT_CNG3 ~ IL_6, data = data), col = "blue", lwd = 2)

plot(data$MCP_1, data$CVLT_CNG3,
     xlab = "MCP-1", ylab = "CVLT Change",
     main = "MCP-1 vs CVLT",
     pch = 19, col = "gray40")
abline(lm(CVLT_CNG3 ~ MCP_1, data = data), col = "blue", lwd = 2)

plot(data$MCP_1, data$CORT_CNG3,
     xlab = "MCP-1", ylab = "Cortical Change",
     main = "MCP-1 vs Cortical",
     pch = 19, col = "gray40")
abline(lm(CORT_CNG3 ~ MCP_1, data = data), col = "blue", lwd = 2)

plot(data$log_IL6, data$CVLT_CNG3,
     xlab = "log(IL-6)", ylab = "CVLT Change",
     main = "log(IL-6) vs CVLT",
     pch = 19, col = "gray40")
abline(lm(CVLT_CNG3 ~ log_IL6, data = data), col = "blue", lwd = 2)

plot(data$log_IL6, data$CORT_CNG3,
     xlab = "log(IL-6)", ylab = "Cortical Change",
     main = "log(IL-6) vs Cortical",
     pch = 19, col = "gray40")
abline(lm(CORT_CNG3 ~ log_IL6, data = data), col = "blue", lwd = 2)

plot(data$log_MCP1, data$CVLT_CNG3,
     xlab = "log(MCP-1)", ylab = "CVLT Change",
     main = "log(MCP-1) vs CVLT",
     pch = 19, col = "gray40")
abline(lm(CVLT_CNG3 ~ log_MCP1, data = data), col = "blue", lwd = 2)

plot(data$log_MCP1, data$CORT_CNG3,
     xlab = "log(MCP-1)", ylab = "Cortical Change",
     main = "log(MCP-1) vs Cortical",
     pch = 19, col = "gray40")
abline(lm(CORT_CNG3 ~ log_MCP1, data = data), col = "blue", lwd = 2)

mtext("Scatterplots of Inflammatory Markers and Outcome Change Scores",
      outer = TRUE, cex = 1.4)

dev.off()

# Reset plotting layout
par(mfrow = c(1, 1))
```


```{r}
# -----------------------
#Correlation table
# -----------------------
# IL-6 vs CVLT
t1 <- cor.test(data$IL_6, data$CVLT_CNG3, use = "complete.obs")
# IL-6 vs Cortical Thickness
t2 <- cor.test(data$IL_6, data$CORT_CNG3, use = "complete.obs")
# MCP-1 vs CVLT
t3 <- cor.test(data$MCP_1, data$CVLT_CNG3, use = "complete.obs")
# MCP-1 vs Cortical Thickness
t4 <- cor.test(data$MCP_1, data$CORT_CNG3, use = "complete.obs")
# IL-6 vs MCP-1
t5 <- cor.test(data$IL_6, data$MCP_1, use = "complete.obs")
# CVLT vs Cortical Thickness
t6 <- cor.test(data$CVLT_CNG3, data$CORT_CNG3, use = "complete.obs")
correlation_table <- data.frame(
  comparison = c(
    "IL_6 vs CVLT_CNG3",
    "IL_6 vs CORT_CNG3",
    "MCP_1 vs CVLT_CNG3",
    "MCP_1 vs CORT_CNG3",
    "IL_6 vs MCP_1",
    "CVLT_CNG3 vs CORT_CNG3"
  ),
  r = c(t1$estimate, t2$estimate, t3$estimate, t4$estimate, t5$estimate, t6$estimate),
  lower = c(t1$conf.int[1], t2$conf.int[1], t3$conf.int[1], t4$conf.int[1], t5$conf.int[1], t6$conf.int[1]),
  upper = c(t1$conf.int[2], t2$conf.int[2], t3$conf.int[2], t4$conf.int[2], t5$conf.int[2], t6$conf.int[2])
)

print(correlation_table)
write.csv(correlation_table, file.path(results_tbl, "correlation_table.csv"), row.names = FALSE)

```
