---
title: "Power_Analysis"
author: "Moj"
date: "2026-03-27"
output: html_document
---

# ============================================================
# BIOS6624 Project 2 - Script 02: Power Analysis
#
# Purpose:
#   - Estimate power for Aim 1 main-effect analyses
#   - Estimate power for Aim 2 interaction analyses
#
# Final strategy:
#   - Aim 1: use corr.1samp
#       * tests whether a marker-outcome correlation differs from 0
#       * one marker at a time, consistent with preliminary data
#
#   - Aim 2: use corr.2samp
#       * tests whether the marker-outcome correlation differs
#         between amyloid groups
#       * amyloid is dichotomized for power calculations only
# Outputs:
#   Results/Tables/aim1_memory_power.csv
#   Results/Tables/aim1_cortical_power.csv
#   Results/Tables/aim2_memory_power.csv
#   Results/Tables/aim2_cortical_power.csv
# ============================================================


```{r cars}

# -----------------------
# Clean environment
# -----------------------
rm(list = ls())

# -----------------------
# Load libraries
# -----------------------
library(powertools)
library(tidyverse)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)


# -----------------------
# Define directories
# -----------------------
results_tbl <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/Results/Tables"
results_fig <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/Results/Figures"
output_dir  <- "C:/Users/mojde/Desktop/Documents/BIOS-6624-Projects/Project2/Output"

# Create folders if they don't exist
dir.create(results_tbl, recursive = TRUE, showWarnings = FALSE)
dir.create(results_fig, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```



```{r cars}
# -----------------------
# Global alpha
# -----------------------
alpha <- 0.05
N_total <- 175

# ============================================================
# Aim 1: Power based on correlation (corr.1samp)
# ============================================================
#
# Rationale:
#   - Preliminary data provide correlation estimates
#   - Power is approximated using one marker at a time tests
#   - This is consistent with the final Aim 1 analysis plan
#   - This also avoids instability from highly correlated markers
#
# Interpretation:
#   - Tests H0: rho = 0
#
# Conservative-scenario logic:
#   - If power is acceptable for the smallest plausible correlation,
#     then power will be greater for the larger correlations
# ============================================================

# -----------------------
# Memory outcome
# -----------------------
# Preliminary data suggested modest associations for memory.
# Conservative scenario is listed first.

aim1_memory_power <- tibble(
  outcome = "Memory",
  scenario = c("Conservative", "Moderate", "Optimistic"),
  rhoA = c(0.20, 0.25, 0.30)
) %>%
  mutate(
    power = map_dbl(
      rhoA,
      ~ corr.1samp(
        N = N_total,
        rho0 = 0,
        rhoA = .x,
        alpha = alpha,
        power = NULL,
        sides = 2,
        v = FALSE
      )
    )
  )

# -----------------------
# Cortical thickness outcome
# -----------------------
# Preliminary data suggested stronger associations for cortical thickness.
# Conservative scenario is listed first.

aim1_cortical_power <- tibble(
  outcome = "Cortical Thickness",
  scenario = c("Conservative", "Moderate", "Optimistic"),
  rhoA = c(0.50, 0.60, 0.65)
) %>%
  mutate(
    power = map_dbl(
      rhoA,
      ~ corr.1samp(
        N = N_total,
        rho0 = 0,
        rhoA = .x,
        alpha = alpha,
        power = NULL,
        sides = 2,
        v = FALSE
      )
    )
  )

print(aim1_memory_power)
print(aim1_cortical_power)

```


```{r cars}
# ============================================================
# Aim 2: Interaction power using correlation comparison
# ============================================================
#
# RATIONALE
# ------------------------------------------------------------
# In the primary analysis, amyloid burden is modeled continuously.
# For power estimation, amyloid is simplified to a binary grouping
# variable:
#
#   A = 0  --> lower amyloid group
#   A = 1  --> higher amyloid group
#
# This allows the interaction term (marker × amyloid) to be
# approximated as a difference in correlations between the two
# amyloid groups:
#
#   rho1 = correlation between inflammation marker and outcome
#          when A = 0 (lower amyloid group)
#
#   rho2 = correlation between inflammation marker and outcome
#          when A = 1 (higher amyloid group)
#
# The null hypothesis is:
#
#   H0: rho1 = rho2
#
# In other words, there is no interaction if the marker-outcome
# association is the same in both amyloid groups.
#
# ------------------------------------------------------------
# WHY USE delta_rho?
# ------------------------------------------------------------
# The interaction effect is represented by the absolute difference:
#
#   delta_rho = |rho2 - rho1|
#
# Here, higher amyloid is hypothesized to strengthen the adverse
# association between inflammation and outcome, so rho2 is set to
# be more negative than rho1:
#
#   rho2 = rho1 - delta_rho
#
# Using delta_rho makes the scenario labels:
#
#   Small     = small interaction effect
#   Moderate  = moderate interaction effect
#   Large     = large interaction effect
#
# ------------------------------------------------------------
# DESIGN NOTE ON GROUP SIZE IMBALANCE
# ------------------------------------------------------------
# Amyloid status is not used to stratify recruitment, so equal
# group sizes cannot be assumed. Because unequal group sizes reduce
# power, we evaluate three plausible allocation scenarios:
#
#   Balanced (50/50)              --> best-case
#   Moderate imbalance (60/40)
#   High imbalance (70/30)        --> conservative case
#
# If power is adequate under the most imbalanced scenario, it will
# be at least as good under more balanced allocations.
# ============================================================


# -----------------------
# Global settings
# -----------------------
N_total <- 175
alpha <- 0.05


# ------------------------------------------------------------
# Group size scenarios
# ------------------------------------------------------------
# n1 = A = 0 group
# n2 = A = 1 group
#
# n_ratio = n2 / n1

ratio_tbl <- tibble(
  ratio_label = c(
    "Balanced (50/50)",
    "Moderate imbalance (60/40)",
    "High imbalance (70/30)"
  ),
  n_ratio = c(1.00, 0.67, 0.43)
) %>%
  mutate(
    n1 = round(N_total / (1 + n_ratio)),
    n2 = N_total - n1
  )


# ------------------------------------------------------------
# Memory outcome
# ------------------------------------------------------------
# Preliminary data suggested modest negative associations between
# inflammatory markers and memory change. To remain conservative,
# the lower amyloid group is assigned a modest baseline correlation.
#
# rho1 = baseline correlation in A = 0
# rho2 = rho1 - delta_rho

aim2_memory_power <- expand_grid(
  outcome = "Memory",
  scenario = c("Small", "Moderate", "Large"),
  ratio_tbl
) %>%
  mutate(
    rho1 = -0.10,
    delta_rho = case_when(
      scenario == "Small" ~ 0.15,
      scenario == "Moderate" ~ 0.25,
      scenario == "Large" ~ 0.35
    ),
    rho2 = rho1 - delta_rho,
    power = pmap_dbl(
      list(n1, n_ratio, rho1, rho2),
      function(n1, n_ratio, rho1, rho2) {
        corr.2samp(
          n1 = n1,
          n.ratio = n_ratio,
          rho1 = rho1,
          rho2 = rho2,
          alpha = alpha,
          power = NULL,
          sides = 2,
          v = FALSE
        )
      }
    )
  ) %>%
  select(
    outcome, scenario, ratio_label,
    n1, n2, n_ratio,
    rho1, rho2, delta_rho, power
  )


# ------------------------------------------------------------
# Cortical thickness outcome
# ------------------------------------------------------------
# Preliminary data suggested stronger negative associations between
# inflammatory markers and cortical thickness than for memory.
# To remain conservative, the lower amyloid group is assigned a
# weaker baseline correlation than the overall pilot estimate.
#
# rho1 = baseline correlation in A = 0
# rho2 = rho1 - delta_rho

aim2_cortical_power <- expand_grid(
  outcome = "Cortical Thickness",
  scenario = c("Small", "Moderate", "Large"),
  ratio_tbl
) %>%
  mutate(
    rho1 = -0.25,
    delta_rho = case_when(
      scenario == "Small" ~ 0.15,
      scenario == "Moderate" ~ 0.25,
      scenario == "Large" ~ 0.35
    ),
    rho2 = rho1 - delta_rho,
    power = pmap_dbl(
      list(n1, n_ratio, rho1, rho2),
      function(n1, n_ratio, rho1, rho2) {
        corr.2samp(
          n1 = n1,
          n.ratio = n_ratio,
          rho1 = rho1,
          rho2 = rho2,
          alpha = alpha,
          power = NULL,
          sides = 2,
          v = FALSE
        )
      }
    )
  ) %>%
  select(
    outcome, scenario, ratio_label,
    n1, n2, n_ratio,
    rho1, rho2, delta_rho, power
  )


# ------------------------------------------------------------
# Combine results
# ------------------------------------------------------------
aim2_power_results <- bind_rows(
  aim2_memory_power,
  aim2_cortical_power
)

print(aim2_power_results)

```



```{r cars}
# ============================================================
# Save tables
# ============================================================

write.csv(
  aim1_memory_power,
  file.path(results_tbl, "aim1_memory_power.csv"),
  row.names = FALSE
)

write.csv(
  aim1_cortical_power,
  file.path(results_tbl, "aim1_cortical_power.csv"),
  row.names = FALSE
)

write.csv(
  aim2_memory_power,
  file.path(results_tbl, "aim2_memory_power.csv"),
  row.names = FALSE
)

write.csv(
  aim2_cortical_power,
  file.path(results_tbl, "aim2_cortical_power.csv"),
  row.names = FALSE
)
```

